15yrsChange

### 2024-04-05 doc/15yrsChange.Rmd
more cleaning

- The future of the critical temperate forest carbon (C) sink is contigent on the role of nuisance species, including non-indigenous pests and pathogens and deer, yet their net effect on current and future C cycling is little understood. Here, we use 15 years of census data from a 25.6-ha plot in a Mid-Atlantic temperate forest to understand how nuisance species are affecting C sequestration. 
- We analyzed biomass change and regenerative capacity in varying sections of the plot classified by deer browsing pressure and the abundance of species vulnerable to non-indigenous pests and pathogens. 
- Biomass mortality increased, driving a net reduction in aboveground C storage, particularly in areas where canopy trees were vulnerable to non-indigenous pests and pathogens. There was also severe depletion of canopy tree species in the understory in areas with high deer browsing, endangering future canopy regeneration.
- Under current trends this forest will continue to lose living biomass and be in danger of a critical transition to a lower biomass ecosystem. The C sequestration of Eastern US deciduous forests is therefore likely overestimated in current global carbon models, and these forests could even transition to a net source of C in the future.

@@

While nuisance species have had a large impact on AGB loss at our site, simultaneous changes including stand aging and climate change have also contributed to observed trends.
The majority of AGB loss has occurred through increased EAB-driven mortality of *Fraxinus spp.* [@anderson-teixeira_longterm_2021], yet mortality among oak species, which is not linked to a non-indigenous nuisance species, is substantial as well (Fig. `r mortality_by_genera`).
As is the case at our site, oak species in the region are generally found in older, relatively even aged cohorts, and this can contribute to accelerated mortality rates (Oak et al., 2016).
However, large oak mortality events, which often occur over protracted periods of time, have been common globally over the past century and are often termed “oak decline” due to the complex abiotic and biotic factors which interact to incite mortality (Manion and Lachance, 1992).
Oak decline is believed to be the result of climate-driven imbalances in carbohydrate dynamics, which weakens oak trees’ resistance to pest and pathogen invasion, eventually leading to mortality (Manion and Lachance, 1992).<!--KAT added "climate-driven" . Not entirely sure that's true -- need to verify.-->
Similarly, stand aging and climate change may both have contributed to the observed decline in AWP (Fig. `r loss_gain_net`b), which in turn contributed to reduced AGB (Figs. `r loss_gain_net`, `r biomass_trends`).
Specifically, reductions in individual tree growth rates at this site are associated with increasing stem diameters, an aging stand, and rising summer temperatures [@dow_warm_2022; @anderson-teixeira_joint_2022].
These complexities underscore the importance of multiple anthropogenic stressors, including but not limited to nuisance species, in shaping 21^st^ century forest carbon dynamics.

@@

The observed trends in AGB loss, including increasing AWM and decreasing AWP, will not necessarily be maintained as continuous directional trends in the future.
The top driver of increasing AWM has been EAB-driven mortality of *Fraxinus spp.*, but *Fraxinus* AWM has started declining as most individuals ≥10 cm have already died (Fig. `r mortality_by_genera`).
It is possible that increasing AWM will be maintained by trends of increasing oak mortality or by other non-indigenous insects or pathogen, which have arrived in the region in multiple waves over the past century [@anderson-teixeira_longterm_2021]. 
The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022].
Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threat_2020]. 
Thus, we consider it likely that oak decline, additional non-indigenous insects or pathogen species, and climate stressors will continue to elevate tree mortality at this site in the coming decades.
Similarly, the observed long-term declines at this site in AWP (Fig. `r loss_gain_net`b) and individual tree growth [@anderson-teixeira_joint_2022] remain poorly understood, hindering prediction of their future trends.
Overall, there remains large uncertainty as to the future trend in AGB at this site, but our analysis makes it clear that these trends have been and will continue to be substantially impacted by nuisance species.

Despite inherent limitations in extrapolating findings into the future and from a single site to a broader region, we find robust evidence suggesting a broader applicability of our results. 
Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region.
Furthermore, the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context.
EAB consistently induces near-total mortality in *Fraxinus* spp. once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list).
Similarly, temperate forests of eastern North America continue to be affected by a variety of non-indigenous pests and pathogens [@anderson-teixeira_longterm_2021; @aukema_historical_2010; @fei_biomass_2019], and emerging threats (e.g., beach leaf disease, sudden oak death) could also have profound impacts [@reed_distribution_2022;@conrad_threat_2020] on forests across the region.
Also paralleling our site, persistently elevated deer populations are reported throughout eastern US forests, and pressures from deer brousing have contributed to regneration debts of canopy species throughout the region [@miller_compounding_2019; @miller_overabundant_2023].

Accounting for nuisance species in carbon models and future forest predictions is imperative. Our results underscore how ignoring nuisance species at our site, and likely much of the region, creates an expectation that temperate decidious forests will keep growing steadily and accumulating C [@wu_uncertainty_2023]. Instead, by increasing tree mortality, nusicance species alter short-term C dynamics and result in large decreases in living woody biomass that are not predicted by models that fail to account for their impact. A result of human introduction and perturbations to ecological equilibria, nuisance species co-occur with other anthropogenic changes and may have interactive effects on forest ecosystems. Through recruitment suppression, nuisance species can delay recovery from disturbances, anthropogenic or natural [@bradshaw_impacts_2016; @harvey_emergent_2023], and may even suppress regeneration into a closed-canopy for sustained periods. Unlike many other disturbances, many non-indigenous pests and pathogens not only induce mortality, but then prevent recovery of their host species [@conrad_threat_2020; @reed_distribution_2022; @knight_factors_2013]. Elimination of canopy species reduces biodiversity, which can decrease resilience to other disturbances (*e.g.* drought; [@anderegg_hydraulic_2018]) further endagerning forest health and C sequestration potential.

@@

All data and code supporting this analysis are available via the public GitHub repository "15yrsChange" within the SCBI-ForestGEO organization (https://github.com/SCBI-ForestGEO/15yrsChange) and archived in Zenodo (DOI: TBD)<!--this can be obtained once the repo is made public-->.
The most up-to-date versions of the data products analyzed here are publicly available via the ForestGEO data portal (https://forestgeo.si.edu/explore-data) and the SCBI-ForestGEO GitHub organization (https://github.com/SCBI-ForestGEO).

>>>>

- The future of the critical temperate forest carbon (C) sink is contingent on the role of nuisance species, including non-indigenous pests and pathogens and deer, yet their net effect on current and future C cycling is little understood. Here, we use 15 years of census data from a 25.6-ha plot in a Mid-Atlantic temperate forest to understand how nuisance species are affecting C sequestration. 
- We analyzed biomass change and regenerative capacity in varying sections of the plot classified by deer browsing pressure and the abundance of species vulnerable to non-indigenous pests and pathogens. 
- Biomass mortality increased, driving a net reduction in aboveground C storage, particularly in areas where canopy trees were vulnerable to non-indigenous pests and pathogens. There was also severe depletion of canopy tree species in the understory in areas with high deer browsing, endangering future canopy regeneration.
- Under current trends this forest will continue to lose living biomass and be in danger of a critical transition to a lower biomass ecosystem. The C sequestration of Eastern US deciduous forests is therefore likely overestimated in current global carbon models, and these forests could even transition to a net source of C in the future.

@@

While nuisance species have had a large impact on AGB loss at our site, simultaneous changes including stand aging and climate change have also contributed to observed trends.
The majority of AGB loss has occurred through increased EAB-driven mortality of *Fraxinus spp.* [@anderson-teixeira_longterm_2021], yet mortality among oak species, which is not linked to a non-indigenous nuisance species, is substantial as well (Fig. `r mortality_by_genera`).
As is the case at our site, oak species in the region are generally found in older, relatively even aged cohorts, and this can contribute to accelerated mortality rates (Oak et al., 2016).
However, large oak mortality events, which often occur over protracted periods of time, have been common globally over the past century and are often termed “oak decline” due to the complex abiotic and biotic factors which interact to incite mortality (Manion and Lachance, 1992).
Oak decline is believed to be the result of climate-driven imbalances in carbohydrate dynamics, which weakens oak trees’ resistance to pest and pathogen invasion, eventually leading to mortality (Manion and Lachance, 1992).<!--KAT added "climate-driven" . Not entirely sure that's true -- need to verify.-->
Similarly, stand aging and climate change may both have contributed to the observed decline in AWP (Fig. `r loss_gain_net`b), which in turn contributed to reduced AGB (Figs. `r loss_gain_net`, `r biomass_trends`).
Specifically, reductions in individual tree growth rates at this site are associated with increasing stem diameters, an aging stand, and rising summer temperatures [@dow_warm_2022a; @anderson-teixeira_joint_2022].
These complexities underscore the importance of multiple anthropogenic stressors, including but not limited to nuisance species, in shaping 21^st^ century forest carbon dynamics.

@@

The observed trends in AGB loss, including increasing AWM and decreasing AWP, will not necessarily be maintained as continuous directional trends in the future.
The top driver of increasing AWM has been EAB-driven mortality of *Fraxinus spp.*, but *Fraxinus* AWM has started declining as most individuals ≥10 cm have already died (Fig. `r mortality_by_genera`).
It is possible that increasing AWM will be maintained by trends of increasing oak mortality or by other non-indigenous insects or pathogen, which have arrived in the region in multiple waves over the past century [@anderson-teixeira_longterm_2021]. 
The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022].
Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threats_2020]. 
Thus, we consider it likely that oak decline, additional non-indigenous insects or pathogen species, and climate stressors will continue to elevate tree mortality at this site in the coming decades.
Similarly, the observed long-term declines at this site in AWP (Fig. `r loss_gain_net`b) and individual tree growth [@anderson-teixeira_joint_2022] remain poorly understood, hindering prediction of their future trends.
Overall, there remains large uncertainty as to the future trend in AGB at this site, but our analysis makes it clear that these trends have been and will continue to be substantially impacted by nuisance species.

Despite inherent limitations in extrapolating findings into the future and from a single site to a broader region, we find robust evidence suggesting a broader applicability of our results. 
Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region.
Furthermore, the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context.
EAB consistently induces near-total mortality in *Fraxinus* spp. once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list).
Similarly, temperate forests of eastern North America continue to be affected by a variety of non-indigenous pests and pathogens [@anderson-teixeira_longterm_2021; @aukema_historical_2010; @fei_biomass_2019], and emerging threats (e.g., beach leaf disease, sudden oak death) could also have profound impacts [@reed_distribution_2022;@conrad_threats_2020] on forests across the region.
Also paralleling our site, persistently elevated deer populations are reported throughout eastern US forests, and pressures from deer brousing have contributed to regneration debts of canopy species throughout the region [@miller_compounding_2019; @miller_overabundant_2023].

Accounting for nuisance species in carbon models and future forest predictions is imperative. Our results underscore how ignoring nuisance species at our site, and likely much of the region, creates an expectation that temperate decidious forests will keep growing steadily and accumulating C [@wu_uncertainty_2023]. Instead, by increasing tree mortality, nusicance species alter short-term C dynamics and result in large decreases in living woody biomass that are not predicted by models that fail to account for their impact. A result of human introduction and perturbations to ecological equilibria, nuisance species co-occur with other anthropogenic changes and may have interactive effects on forest ecosystems. Through recruitment suppression, nuisance species can delay recovery from disturbances, anthropogenic or natural [@bradshaw_impacts_2016; @harvey_emergent_2023], and may even suppress regeneration into a closed-canopy for sustained periods. Unlike many other disturbances, many non-indigenous pests and pathogens not only induce mortality, but then prevent recovery of their host species [@conrad_threats_2020; @reed_distribution_2022; @knight_factors_2013]. Elimination of canopy species reduces biodiversity, which can decrease resilience to other disturbances (*e.g.* drought; [@anderegg_hydraulic_2018]) further endagerning forest health and C sequestration potential.

@@

All data and code supporting this analysis are available via the public GitHub repository "scbi_nuisance_species_biomass" within the SCBI-ForestGEO organization (https://github.com/SCBI-ForestGEO/scbi_nuisance_species_biomass). <!-- and archived in Zenodo (DOI: TBD)this can be obtained once the repo is made public-->
The most up-to-date versions of the data products analyzed here are publicly available via the ForestGEO data portal (https://forestgeo.si.edu/explore-data) and the SCBI-ForestGEO GitHub organization (https://github.com/SCBI-ForestGEO).
